---
title: "Package_Details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package_Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fatherStay)
```

## Objective

Our data are expected to look something like the following:

| date | y   | x1  | z1  | z2  |
|------|-----|-----|-----|-----|
| 1    | 1.1 | 0.5 | 2.3 | 1.2 |
| 2    | 1.5 | 0.2 | 2.7 | 1.2 |
| 3    | NA  | 0.1 | 2.8 | 1.5 |
| 4    | 1.5 | NA  | 2.7 | 1.3 |
| 5    | 1.8 | 0.2 | 2.9 | 1.2 |
| ...  | ... | ... | ... | ... |
| 204  | 2.5 | 1.2 | 1.7 | 2.2 |
| 205  | 2.9 | 0.8 | 0.7 | 2.5 |
| 206  | NA  | 0.2 | 2.7 | 2.9 |
| 207  | NA  | 0.6 | 2.2 | 1.9 |
| 209  | NA  | 0.4 | 2.0 | 3.9 |
| 210  | NA  | 0.2 | 1.0 | 3.8 |

Note the string of NA values in y in the last days of the study, which are the values to be nowcasted. Note also the presence of NA values in the response and the predictors.

## Imputation

We expect there to be some missing data in the predictor variables. Imputation is the best way to deal with them. The goal is that we will have relatively simple methods to impute the values of the explanatory variables $x_1$, $z_1$, and $z_2$, and the package will allow the user to specify the type of imputation to be used. Additionally our data has the advantage of having multiple potentially correlated variables. We would ideally like to use this fact to our advantage when imputing the data to create better imputations. Imputation is not the main focus of this project and so we will aim to keep the imputation methods used relatively simple. 

There are several existing R packages that allow for imputation and work in a time series context, and while we do not want the program we are developing to be dependent on many packages it makes sense to use an existing package for the imputation since it is not a focus of the project. 

The package that we will be using in this project is imputeTS, a package in R for imputing univariate time series. It has the advantage of being specifically for time series data. It features a variety of complex methods for imputation such as; Kalman smoothing, and interpolation, as well as simple methods like using the last observation or the mean. While this package does not take advantage of the multivariate nature of our data it is still a good option to impute the missing values in our data.

## Nowcasting

Starting with defining the specific problem that we are wanting to solve. When referring to nowcasting we denote the measurement of DAD at time $t$ by the value $Y_t$ and the measurements of the explanatory variables at time $t$ to be the values $X_{1,t}$ for RVDSS and $Z_{j,t}$ where $j \in \{1,2\}$ representing CNISP, and PTSOS. Then the problem that we are working is how to predict the value of $Y_t$ given the values $X_{1,t}$, $Z_{j,t}$, and $Y_{t-1}$.

We will be making several key assumptions in this project. The first is that the data for the explanatory variables can be at least somewhat successfully imputed by reasonable estimates. We will also assume that we have access to all the updated values of the explanatory variables at the time $t$ when we are making a prediction for $Y_t$. <- Devan, is this a good/valid assumption to make?

There are several different techniques for nowcasting in a time series context, in general they are organized from simplest to more complex:

### Simple Linear Models

The first and simplest models that we will be looking at in this project are linear models where the value of DAD we are nowcasting is dependent on the current values of the explanatory variables:
$$
y_t = ax_t + bz_{t,i} + \epsilon_t
$$

We can build on this basic model by adding the complexity of lagging the values of the explanatory variables:
$$
y_t = cx_t + dx_{t-1} + \epsilon_t
$$
Building on this complexity is if we take into account previous iterations of $y$, which is effectively a simple AR model of $y$ based on $x$:
$$
y_t = ey_{t-1} + fx_{t} + \epsilon_t
$$

### Non Linear Models
In additional all of these relatively simple linear models we will also be considering non-linear models/processes. In this project we are seeking to predict the current values of DAD, not create an explanatory model. This makes many ML models a good option to use since they are focused on creating predictions not on their explanatory power. We will start with using random forests, since they are relatively easy to implement and are a powerful tool, the predictive capabilities can be further enhanced by using techniques like XGboost. 

The models discussed here are the ones we will definitely be using and a good starting point, but throughout the project more will likely come up and we will be using those as well. We are not limiting ourselves to just the models discussed here.

### ARX

A natural next step is to model the process as an ARX process, that is as an AR system that is also dependent on current and past values of some explanatory variable. We can represent it by the following equations:

$$
y_t = a_{t} x_{t} + \sum_{i=1}^n a_{t-i} x_{t-i} + \sum_{j=1}^m b_{t-i} y_{t-i} + \epsilon_t
$$
This system can also be expanded to incorporate multiple explanatory variables:

$$
y_t = a_{t} x_{t} + \sum_{i=1}^n a_{t-i} x_{t-i} + \sum_{j=1}^m b_{t-i} y_{t-i} + \sum_{i=1}^p c_{t-i} z_{t-i,1} + c_{t} z_{t,1} +  \epsilon_t
$$
It would be easier to represent this in a vector form as:

$$
y_t = a_{t} x_{t} + \boldsymbol{a}\boldsymbol{x} + \boldsymbol{b}
\boldsymbol{y} + \boldsymbol{c} \boldsymbol{z_1}  + c_{t} z_{t,1} + \epsilon_t
$$

Where $\boldsymbol{x, \ y, \ z}$ are the past values of those variables, and $\boldsymbol{a, \ b, \ c}$ are the values of the coefficients.

### Kalman filter/Smoothing

A common method in nowcasting time series data is using the Kalman Filter (rossiKalmanFilteringBased2015). The Kalman filter predicts one step ahead and its algorithm works in two steps, first creating a prediction, and then after the actual data is observed, updating that prediction using the Kalman gain. This preserves the overall trend of the data, while also taking the actual values into account. The applications of this to nowcasting are that we would use the prediction step to nowcast the desired value (DAD) and then the prediction step would update the algorithm when more data is observed.

A Kalman filter is capable of handling the multivariate nature of this data. The foundational equations that the Kalman filter is based on:
$$
y_t = a_tx_t + \epsilon_{1,t} \\
x_t = b_tx_{t-1} + \epsilon_{2,t}
$$

Can easily be changed to accommodate multiple predictors:

$$
y_t = a_tx_t + b_tz_t + \epsilon_{1,t} \\
x_t = c_tx_{t-1} + \epsilon_{2,t} \\
z_t = d_tz_{t-1} + \epsilon_{3,t}
$$

There are R packages that can be used to run a Kalman Filter such as kalmanfilter these are a good option for us to use since manually coding a Kalman filter would be very complex.
